---
output:
  html_document:
    df_print: paged
---

# Data-driven identification of US communities at additional risk for COVID-19 related health conditions

- Andrew Deonarine, Jerod Parrent, Chirag J Patel, Arjun K Manrai 
- See our blog post: https://www.xy.ai/process/covid-19-communities-at-risk
- And second post: https://www.xy.ai/process/covid-19-are-we-testing-the-highest-risk-communities
- last updated: 3/17/2020


# Data sources used below

- COVID-19 tracking: https://covidtracking.com/api/
- CDC 500 Cities (2016-2017 estimates): https://chronicdata.cdc.gov/browse?category=500+Cities
- XY Exposome Data Warehouse, US Census estimated (2015 5-year estimate): https://www.census.gov/programs-surveys/acs


```{r library load, echo=FALSE}
# load libraries
library(tidyverse)
library(ggthemes)
library(ggrepel)
library(forcats)
library(readr)
library(DT)
```


# Query the Current Covid Cases in the United States
```{r, echo=TRUE}
# from covidtracking.com API
#current_covid <- read_csv(url('http://covidtracking.com/api/states.csv')) #http://covidtracking.com/api/states.csv
current_covid <- read_csv('./states.csv')
current_covid <- current_covid %>% mutate(state = fct_reorder(state, (positive)))
p <- ggplot(current_covid, aes(state, positive))
p <- p + geom_point() + coord_flip()
p <- p + theme_fivethirtyeight() + theme(axis.title = element_text(), legend.position = 'none', axis.text = element_text(size=8)) + labs(y='Positive Cases', x='State')
p

p <- ggplot(current_covid, aes((positive+negative), I((positive/(positive+negative))*100), label=state))
p <- p + geom_text(size=3) + ylab('Fraction Positive') + xlab('Total Tested') + scale_x_log10()
p <- p + theme_fivethirtyeight() + theme(axis.title = element_text(), legend.position = 'none') + labs(x='Total Tested', y='Percent Positive of Those Tested')
p

datatable(current_covid %>% arrange(desc(positive)))

```


## Load in the 500 Cities Data, merged with EDW on a census-tract level
```{r get data from EDW, echo=TRUE}
fh_cities <- read_csv('fh_cities_census.csv')
head(fh_cities)
```


## Community Comorbidity Risk Score: a function of Cancer, Stroke, Asthma, COPD, CHD, CKD, Obesity, Smoking, and Diabetes

```{r, echo=TRUE}

#fh_cities <- fh_cities %>% mutate(comorbidity_risk_score = (CANCER_CrudePrev + STROKE_CrudePrev + CASTHMA_CrudePrev + COPD_CrudePrev + CHD_CrudePrev + OBESITY_CrudePrev + CSMOKING_CrudePrev + DIABETES_CrudePrev))

fh_cities <- fh_cities %>% filter(population_2010 > median(fh_cities$population_2010)) # filter by high population census tracts
fh_cities <- fh_cities %>% mutate(comorbidity_risk_score = (scale(CANCER_CrudePrev) + scale(STROKE_CrudePrev) + scale(CASTHMA_CrudePrev) + scale(COPD_CrudePrev) + scale(CHD_CrudePrev) + scale(OBESITY_CrudePrev) + scale(CSMOKING_CrudePrev) + scale(DIABETES_CrudePrev) + scale(KIDNEY_CrudePrev)))


fh_cities$comorbidity_risk_score <- scale(fh_cities$comorbidity_risk_score) # z-score the risk score
fh_cities <- fh_cities %>% filter(!is.na(median_age)) 

# top census tracts per state, but filtered for those that have a high risk score
top_per_state <- fh_cities %>% group_by(stateabbr) %>% top_n(1,comorbidity_risk_score) %>% ungroup() %>% filter(comorbidity_risk_score >= 2.5)
# top census tracts per state based on age, but filtered for those that have a high risk score
oldest_tracts <- fh_cities %>% group_by(stateabbr) %>% top_n(1,median_age) %>% ungroup() %>% filter(median_age >= 50)
top_both <- fh_cities %>% filter(comorbidity_risk_score > 2 & median_age >= 55)

### main plot 
p <- ggplot(fh_cities, aes(median_age, comorbidity_risk_score))
p <- p + geom_point(alpha=0.5, color='gray')

p <- p + geom_point(data=top_per_state, aes(median_age, comorbidity_risk_score, size=population_2010))
p <- p + geom_text_repel(data=top_per_state, aes(median_age, comorbidity_risk_score, label=paste(placename, stateabbr)), size=3)

p <- p + geom_point(data=oldest_tracts, aes(median_age, comorbidity_risk_score, size=population_2010), color='red')
p <- p + geom_text_repel(data=oldest_tracts, aes(median_age, comorbidity_risk_score, label=paste(placename, stateabbr)),color='red', size=3)


p <- p + theme_fivethirtyeight() + theme(axis.title = element_text(), legend.position = 'none') + labs(x = 'Median Age of Tract', y = 'Comorbidity Risk Score')
p


```

## Comorbidity Risk Score and Lack of Access To Care
```{r, echo=TRUE}
lack_access <- fh_cities %>% filter(comorbidity_risk_score >= 3, ACCESS2_CrudePrev > 25)
p <- ggplot(fh_cities, aes(ACCESS2_CrudePrev, comorbidity_risk_score, size=population_2010))
p <- p + geom_point(alpha=0.5, color='gray')
p <- p + geom_point(data=lack_access, aes(ACCESS2_CrudePrev, comorbidity_risk_score), color='black')
p <- p + geom_text_repel(data=lack_access, aes(ACCESS2_CrudePrev, comorbidity_risk_score, label=paste(placename, stateabbr)),color='black ', size=3)
p <- p + theme_fivethirtyeight() + theme(axis.title = element_text(), legend.position = 'none') + labs(x = '% Lack of access to Healthcare', y = 'Comorbidity Risk Score')
p
```


## Comorbidity Risk Score, Age, and High Proportion of Positive Cases
```{r, echo=TRUE}
fh_covid_cases <- fh_cities %>% left_join(current_covid, by=c('stateabbr'='state')) %>% mutate(covid_percent = positive / (positive+negative))
top3_como_covid <- fh_covid_cases %>% group_by(covid_percent) %>% top_n(3, comorbidity_risk_score) %>% ungroup() %>% filter(covid_percent >= .15, total >= 100)
p <- ggplot(fh_covid_cases %>% filter(covid_percent >= .15, total >= 100), aes(covid_percent*100, comorbidity_risk_score, size=population_2010))
p <- p + geom_jitter(alpha=0.5, color='pink')
p <- p + geom_text_repel(data=top3_como_covid, aes(covid_percent*100, comorbidity_risk_score, label=placename))
p <- p + geom_label_repel(data=top3_como_covid %>% group_by(stateabbr) %>% top_n(1, comorbidity_risk_score), aes(covid_percent*100, -2, label=stateabbr), size=3)
p <- p + theme_fivethirtyeight() + theme(axis.title = element_text(), legend.position = 'none') + labs(x = 'Percent tested positive in state [min tests of 100]', y = 'Comorbidity Risk Score')
p

top3_age_covid <- fh_covid_cases %>% group_by(covid_percent) %>% top_n(3, median_age) %>% ungroup() %>% filter(covid_percent >= .15, total >= 100)
p <- ggplot(fh_covid_cases %>% filter(covid_percent >= .15, total >= 100), aes(covid_percent*100, median_age, size=population_2010))
p <- p + geom_jitter(alpha=0.5, color='pink') 
p <- p + geom_text_repel(data=top3_age_covid, aes(covid_percent*100, median_age, label=placename))
p <- p + geom_label_repel(data=top3_age_covid %>% group_by(stateabbr) %>% top_n(1, median_age), aes(covid_percent*100, 20, label=stateabbr), size=3)
p <- p + theme_fivethirtyeight() + theme(axis.title = element_text(), legend.position = 'none') + labs(x = 'Percent tested positive in state [min tests of 100]', y = 'Median Age')
p

```


## Senstivity Analyses 

### Examine the top community/census tract for specific diseases, such as Cancer, Diabetes, Asthma, and CHD
```{r fig.width=12, fig.height=12}
# first scale all
fh_covid_cases_sens <- fh_covid_cases %>% mutate(CANCER_CrudePrev=scale(CANCER_CrudePrev), STROKE_CrudePrev=scale(STROKE_CrudePrev), CASTHMA_CrudePrev=scale(CASTHMA_CrudePrev), COPD_CrudePrev=scale(COPD_CrudePrev), CHD_CrudePrev=scale(CHD_CrudePrev), OBESITY_CrudePrev=scale(OBESITY_CrudePrev), CSMOKING_CrudePrev=scale(CSMOKING_CrudePrev), DIABETES_CrudePrev=scale(DIABETES_CrudePrev), KIDNEY_CrudePrev=scale(KIDNEY_CrudePrev))


# show Asthma, CHD, T2D, cancer
fh_covid_long <- fh_covid_cases_sens[, c('fips_place_tract',  'CANCER_CrudePrev', 'DIABETES_CrudePrev', 'CASTHMA_CrudePrev', 'CHD_CrudePrev')] %>% gather('disease', 'score', -fips_place_tract)

fh_covid_long <- fh_covid_long %>% left_join(fh_covid_cases_sens %>% select(fips_place_tract, placename, stateabbr, population_2010, covid_percent, total) %>% unique(), by='fips_place_tract')

fh_covid_long$disease <- tolower(unlist(lapply(strsplit(fh_covid_long$disease, "_"), function(arr) {arr[[1]]})))

top_disease_covid <- fh_covid_long %>% group_by(covid_percent, disease) %>% top_n(1, score) %>% ungroup() %>% filter(covid_percent >= .15, total >= 100)

p <- p <- ggplot(fh_covid_long %>% filter(covid_percent >= .15, total >= 100), aes(covid_percent*100, score, size=population_2010))
p <- p + geom_jitter(alpha=0.5, color='pink')
p <- p + geom_label_repel(data=fh_covid_long %>% group_by(stateabbr, disease) %>% top_n(1, score) %>% ungroup() %>% filter(covid_percent >= .15, total >= 100), aes(covid_percent*100, 0, label=stateabbr), size=3)
p <- p + facet_wrap(.~disease, nrow=2)
p <- p + geom_text_repel(data=top_disease_covid, aes(covid_percent*100, score, label=placename))
p <- p + theme_fivethirtyeight() + theme(axis.title = element_text(), legend.position = 'none') + labs(x = 'Percent tested positive in state [min tests of 100]', y = 'Disease-Specific Risk Score')
p

```

We examine the similarities between the COVID-19 risk communities and specific disease: 
- a community in Shreveport ranks high for all of these candidate diseases, such including cancer, asthma, heart disease (chd in the figure), and diabetes. We conclude that Shreveport is a community of concern. 

Differences:
- The COVID-19 risk score for a community in Pembroke Pines, FL seems to be dominated by heart disease
- The COVID-19 risk score for a community in Akron, OH and Camden, NJ seems to be dominated by diabetes



## News and social media updates on at-risk populations:

- https://www.nytimes.com/2020/03/12/health/coronavirus-midlife-conditions.html
- https://twitter.com/search?q=%23HighRiskCovid19&src=trend_click
